---
title: "Survival Statistics"
author: ''
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
params: 
  data: "test/worms1.csv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
## load libraries
packages = c('knitr','reshape2','tidyr','survival','survminer','dplyr','MASS','stringr')
# check.packages function: install and load multiple R packages.
# Check to see if packages are installed. Install them if they are not, then load them into the R session.
check.packages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}
msg.trap <- capture.output( suppressMessages( check.packages(packages)) ) # silent mode
```

## Data

Raw data:

```{r raw_data}
## Read data from usual dataset (day, condition1, condition2, etc.)
worms <- read.csv(params$data, header = TRUE)
knitr::kable(worms)
```

The survival data was reformatted so the Kaplan-Meier plot could be produced. Each row represents a different test subject. If the test subject survived the whole study `fustat = 0` but if the subject died before the end of the study `fustat = 1`. `futime` reflects the day/time subject had died or, if the subject survived the study, `futime` reflects the day/time when the study ended. Tick marks indicate that one or more subjects have been excluded from the study at a particular time point, or (if the tick mark appears on the final time point) that the study has been concluded. Below is an example of the first 5 rows of data. The full transformed data set can be found in the output file `FormattedSurvivalData.csv`. 

```{r k-m_data}
## import data so it no longer has header (easier for parsing)
worms <- read.csv(params$data, header = FALSE)

## store important info as variables
time_type = worms[1,1]
time_interval_v = as.numeric(as.vector(worms[[1]])[-1])
starting_worms_v = as.numeric(worms[2,][-1])
conditions_v = as.character(worms[1,][-1])
study_length = as.numeric(nrow(worms))-2
total_worms = sum(starting_worms_v)

## format intermediate data frame for downstream use
worms.t <- data.frame(t(worms[-1])) # transpose original dataframe
colnames(worms.t) = worms[, 1] # rename columns
names(worms.t)[1] = "condition"
worms.melted <- melt(worms.t, id.vars=c('condition')) # melt dataframe
names(worms.melted)[2] = time_type
names(worms.melted)[3] = "live_count"
subtr = rep(starting_worms_v, times = study_length+1)
worms.melted$subtr = subtr
worms.melted$dead_count = worms.melted$subtr - as.numeric(worms.melted$live_count)
worms.melted <- with(worms.melted,  worms.melted[order(worms.melted$condition) , ])
row.names(worms.melted) <- 1:nrow(worms.melted) # renumber rows after reordering

## Calculate fustat & futime
##   fustat = 1 if the worm died before the end of the study
##   fustat = 0 if the worm survived the whole study
##   futime = time worm died (if fustat = 1), or last day of study (if fustat = 0)

futime = as.vector(c())
fustat = as.vector(c())
worms.melted = na.omit(worms.melted)
for (k in seq(from=0, to=as.numeric(length(conditions_v)*study_length), 
              by=study_length+1)) {
  daily_death_v = as.vector(c())
  for (i in 1:study_length){
    daily_death = as.vector(worms.melted$dead_count[k+i+1] - worms.melted$dead_count[k+i])
    daily_death_v = append(daily_death_v, daily_death)
  }
  ending_dead_worms = sum(daily_death_v)
  ending_live_worms = as.numeric(worms.melted$live_count[k+1]) - ending_dead_worms
  fustat = append(fustat, rep(1, times=ending_dead_worms))
  fustat = append(fustat, rep(0, times=ending_live_worms))
  futime = append(futime, rep(time_interval_v[-1], times=daily_death_v))
  futime = append(futime, rep(study_length, times=ending_live_worms))
}

## Build new data frame for stats
condition = sort(rep(as.vector(worms.t[[1]]), times = starting_worms_v))
worm.df = data.frame(condition)
worm.df$fustat = fustat
worm.df$futime = futime

## write formatted data to file
write.csv(worm.df, file="FormattedSurvivalData.csv", row.names = TRUE)

## Show new formatted survival dataset and output csv
kable(worm.df[1:5,], row.names = TRUE)
```

## Kaplan-Meier Plot

This plot shows the fraction of surviving test subjects from each treatment group by each day of treatment. 

```{r k-m_plot}
time_gap = time_interval_v[2] # determines breaks on x-axis
surv_object <- Surv(time = worm.df$futime, event = worm.df$fustat)
fit1 <- do.call(survfit, list(surv_object ~ condition, data = worm.df))
ggsurvplot(fit1, data = worm.df, xlim = c(0,study_length), 
           break.time.by = time_gap, 
           xlab = paste("Time (",time_type,")",sep=""), legend="right") 
```

## Cox Proportional-Hazards Model 

This plot reflects the risk of test subjects dying during the specified time interval. The higher the ratio value, the more likely the subjects in that treatment group will die (compared to a control group). The reference group is indicated as the first group listed in this figure. 

```{r cox}
fit.coxph <- coxph(surv_object ~ condition, data = worm.df)
ggforest(fit.coxph, data = worm.df)
```

## Log Rank Tests

Log rank tests compare whether the results from a pair of test conditions are significantly different from one another. If the p-value is < 0.05, the two test conditions are significantly different. 

```{r pressure}
cmbo = data.frame(t(combn(conditions_v,2)))
num_comb = as.numeric(nrow(cmbo))
surv_pval_v = as.vector(c())
for (i in 1:num_comb){
  subset_pair = worm.df[which(worm.df$condition==cmbo[i,1]
                              |worm.df$condition==cmbo[i,2]),]
  surv_object <- Surv(time = subset_pair$futime, event = subset_pair$fustat)
  fit1 <- do.call(survfit, list(surv_object ~ condition, data = subset_pair))
  surv_curve <- ggsurvplot(fit1, pval = TRUE, data = subset_pair, xlim = c(0,study_length), 
                           break.time.by = time_gap, 
                           xlab = paste("Time (",time_type,")",sep=""), legend="right")  
  surv_pval = surv_pvalue(fit1, data=subset_pair)$pval
  surv_pval_v = append(surv_pval_v, surv_pval)
}
cmbo$pval = surv_pval_v
colnames(cmbo) <- c('condition1','condition2','pvalue')

## Show Log Rank Values
kable(cmbo)
```

## LT50 (Lethal Time)

The LT50 value is the time at which 50% of subjects are dead. 

```{r LT50}
## Calculate LT50
lt50.df = data.frame(conditions_v)
lt50_v = as.vector(c())
day.list = c(1:(study_length+1)) #this makes day 0 = day 1, so subtract 1 from Dose, below
for (i in 1:as.numeric(length(conditions_v))){
  subset = worms.melted[which(worms.melted$condition==conditions_v[i]),]
  subset$live_count = as.numeric(subset$live_count)
  y = cbind(subset$live_count,subset$dead_count)
  model.results = glm(y~day.list,binomial)
  lt50 = dose.p(model.results,p=0.5)[1]-1 # what time gives 50% survival
  lt50 = as.numeric(round(lt50, digits=1))
  lt50_v = append(lt50_v, lt50)
}
lt50.df$lt50 = lt50_v
colnames(lt50.df) = c('condition','LT50')

## Output csv for LT50 (Lethal Time)
kable(lt50.df)
```


